Parameter sensitivity and identifiability for a biogeochemical model of hypoxia in the northern {G}ulf of {M}exico\tnoteref{mytitlenote}

### 2017-08-01 ident_manu.tex
revisions half way done

\title{Parameter sensitivity and identifiability for a biogeochemical model of hypoxia in the northern {G}ulf of {M}exico\tnoteref{mytitlenote}}
\tnotetext[mytitlenote]{Version: Tue May 2 09:04:03 2017 -0500, \href{https://github.com/fawda123/identifiability/commit/eedd4b1c3c2a4bd984b414a5b3fc28e9845d2d88}{eedd4b1c3c2a4bd984b414a5b3fc28e9845d2d88}\newline Acronyms: \acfi{chla}; \acfi{cgem}; \acfi{do}; \acfi{dom}; \acfi{gom}; \acfi{lcs}; \acfi{marb}; \acfi{pom}; \acfi{rmse}; \acfi{zerod}}

@@

\begin{abstract}
\noindent Local sensitivity analyses and identifiable parameter subsets were used to describe numerical constraints of a hypoxia model for bottom waters of the northern Gulf of Mexico.  The sensitivity of state variables differed considerably with parameter changes, although most variables were responsive to changes in parameters that influenced planktonic growth rates and less sensitive to physical or chemical parameters.  Variation in sensitivity had a direct correspondence with identifiability, such that only small subsets of the complete parameter set had unique effects on the model output. Selecting parameters by decreasing sensitivity demonstrated that only eight of 51 total parameters had a sufficiently unique effect on model output for precise calibration.  As a result, parameter selection heuristics were used to identify parameters for model calibration that depended on combined effects on output, relative sensitivity of each parameter, and ecological categories for the biogeochemical equations. The calibrated \ac{zerod} unit of the hypoxia model had improved precision if sensitive parameters were included in an identifiable subset.  Extension of results to a three-dimensional grid of the Gulf of Mexico showed that sensitive parameters for the \ac{zerod} model translated to non-trivial changes in the areal estimates of hypoxia.
\end{abstract}

@@

Calibration of parameter values for each of the seven selection heuristics (\cref{tab:heurist1,tab:heurist2,tab:heurist3})  produced model output that varied in precision and computational effort required to identify optimal values (\cref{tab:calib}).  As expected, the number of required iterations of the optimization algorithm to minimize \ac{rmse} of observed \ac{do} and model output varied in proportion to the number of parameters in a subset.  For example, only six iterations were required to optimize the single parameter selected for the temperature category, whereas 1530 iterations were required for the eight parameters selected independent of category.  No association was observed between collinearity ($\gamma$) and number of iterations or calibration.  Of the seven parameter subsets, selection independent of parameter category and equally within each category produced the largest reduction in \ac{rmse} values from the default output (starting \ac{rmse} $72.14$, reduction to $66.77$, $67.44$, respectively), whereas the optimal values for parameters in the Optics category had the smallest reduction ($72.12$).  Lower \ac{rmse} was associated with higher average sensitivity of a parameter subset.  Parameters selected only in the optics category had the highest \ac{rmse} and lowest average L1 ($2.81\times 10^{-4}$, three parameters), whereas parameters selected independent of category had the lowest \ac{rmse} and highest average L1 ($0.03$, eight parameters, \cref{tab:calib}).

@@

Common goals in the application of biogeochemical models are to \begin{inparaenum}[1\upshape)]
\item accurately describe the system by matching predictions with observed data \citep{Reckhow90}, and 
\item provide a means of forecasting ecosystem condition with hypothesized management or environmental scenarios \citep{Clark01}. 
\end{inparaenum}
Although these objectives are the focus of most applications, the structural components and associated parameters of process-based models should secondarily provide insight of ecosystem processes that are driving observed changes.  This latter objective represents a more generic scientific purpose of modelling that extends beyond the benefits of describing and predicting change.  Modelers aim to identify universal principles that govern dynamics across systems and the constraint of model parameters to observations provides a means of supporting or refuting hypotheses \citep{Hilborn97,Kirchner06}.  Accordingly, this study provided a simple approach to use the effects of parameter perturbations on state variables to characterize identifiable parameter subsets that varied by selection criteria.  By doing so, we provided a simple and preliminary approach towards improved model refinement that begins to address more general purposes of model applications. Our specific results demonstrated that small parameter subsets relative to all sensitive parameters were within the identifiability thresholds described in the literature.  The identifiable parameter subsets varied considerably between state variables and the method for parameter selection.  Use of identifiable parameter subsets with sensitive parameters improved model precision and results for the smaller model were scalable to the 3-dimensional model.  In general, these results provide justification for the use of explicit parameter selection heuristics to improve model precision.

State variables were most sensitive to parameters in the phytoplankton and zooplankton categories, particularly the maximum growth rates (\textit{umax} for phytoplankton, \textit{Zumax} for zooplankton), mortality coefficient for phytoplankton (\textit{mA}), and the zooplankton half saturation coefficent for grazing (\textit{ZKa}). An increase in the growth rate of primary producers has the potential to increase oxygen concentration through photosynthetic processes, although increased production of organic matter is balanced with respiration and bacterial decomposition that reduce \ac{do} in the water column.  Similarly, increases in zooplankton abundance with increased growth rates causes a reduction in phytoplankton biomass through grazing, which is expected to further deplete pools of organic matter from algal sources.  Most variables were also sensitive to variation in the half-saturation grazing coefficient which moderates the concentration of nutrients that support half the maximum grazing rate. Although the tradeoff between abundance, grazing, and decomposition is complex, the sensitivity of model state variables to parameters that directly control the abundance of primary producers is in agreement with empirical observations of factors that influence hypoxia dynamics on the \ac{lcs} \citep{Fahnenstiel95,Roelke00,Eldridge10}. The sensitivity of the model output to variation in other parameters that relate to physical and chemical properties of the system was of secondary importance to biological relationships.  That is, state variables were sensitive to changes in light and temperature parameters, although to a lesser extent than phytoplankton and zooplankton parameters.  As such, the differing sensitivities of state variables to parameters in each of the categories was not unexpected given general ecological relationships that are well understood and described by the model.

A general conclusion from the identifiability analyses is that only limited subsets of parameters were identifiable within the constraints of local sensitivity analyses. These results support previous studies that have suggested similarly small subsets of parameters can be identified using traditional calibration schemes \citep[e.g.,][]{Wheater86,Ye97,Omlin01}.  In addition to \ac{cgem}, these conclusions have relevance for other biogeochemical models that include numerous parameters and structural equations to characterize processes in the model domain.  A general conclusion is that less complex models could potentially be beneficial given that only a small subset of parameters is identifiable and that ecosystem processes may in fact be sufficiently characterized with few parameters \citep{Ye97}.  Conversely, others have argued that model complexity is not in itself a disadvantage when parsimony is not the sole arbitrator of model structure \citep{Reichert97}. Over-parameterization can be useful if processes have importance that were not evaluated during model identification.  Single objective functions that maximize model precision with identifiable parameters may also provide an incomplete characterization of model worth, which has prompted the development of probability-based models of hypoxia that explicitly include uncertainty in model components \citep[e.g.][]{Obenour15}.  Our results demonstrated that approximately 75\% of the evaluated parameters had an effect on the eight state variables, whereas \ac{cgem} includes a total of 36 variables and multiple plankton groups, not all of which have immediate concern for understanding hypoxia. The redundancies identified with the sensitivity analyses are only problematic if the primary interest is, for example, \ac{do} dynamics.  Moreover, the proposed selection heuristics provide flexibility for choosing different parameters with the assumption that those chosen depend on the research or management question.

Results from the identifiability analyses provided additional insight into the interactions of parameters in large biogeochemical models.  First, identifiability of parameter subsets was not related to the sensitivity of individual variables. As noted above, an identifiable parameter is one that has a unique effect on model predictions that cannot be compensated for or undone by changing other parameters. The magnitude of the effect of a parameter has no bearing on identifiability, which further complicates the selection of parameters for calibration.  Although identifiability is the primary limiting factor in choosing a set, the relative sensitivities are more important for the decision to include or exclude individual parameters.  For example, \cref{tab:calib,fig:areachg} demonstrated that parameter sets with sensitive parameters had improved model precision and that these parameters had affects that translated to the larger model. Our analysis addressed this challenge by presenting multiple selection criteria for identifiable parameter sets that prioritized the most sensitive parameters during the selection process.  Similarly, identifiability was not always related to the number of parameters in a set. Although the overwhelming trend was decreasing identifiability with more parameters, the unique effects of including an individual parameter with an existing set often reduced the $\gamma$ estimate. For example, \cref{fig:exclex} showed that including \textit{KGcdom}, \textit{KO2}, or \textit{nitmax} in parameter sets more often reduced $\gamma$ relative to sets that excluded the parameters.  These examples demonstrate the complex interactions of parameter changes on variable response, highlighting the need to consider the combined and individual effects of parameters on identifiability. The selection criteria proposed in our above analyses can facilitate parameter selection and also provide diagnostic tools to identify parameters with disproportionate effects on $\gamma$.

@@

An evaluation of sensitivity and identifiability of relevant parameter sets is a preliminary and simplistic approach to improving model predictions.  In general, uncertainty analyses that lead to improved models are ultimately expected to increase our understanding of properties defining ecological relationships.  The extension of simple parameter sensitivity analyses to the generalization of ecosystem properties requires additional analysis and model refinement. Further, the utility of a specific model depends on the question and objective for application to a specific system.  For the above analysis, the FishTank model, as part of the larger \ac{cgem} application, was evaluated in the context of hypoxia effects on ecosystem condition and function.  Our results have shown that relatively small subsets of parameters are identifiable given the complexity of the model, and as a result, we have provided a general approach to select parameter subsets depending on the ecological context (i.e., selection by parameter category, selection for specific state variables). We further demonstrated the effects of selected subsets on model precision and the potential for scalability to larger spatial-temporal models. Thus, the results described above have relevance for further model refinement with the specific goal of better understanding ecological dynamics that moderate hypoxia on the northern \ac{gom}.  However, the general principles of sensitivity and parameter identifiability have broad applicability beyond this context and we argue that such methods should be more universally applied as an initial approach to quantify numerical constraints of biogeochemical models.

Specific approaches can be used to improve and build on the results presented herein, in addition to the more general considerations above.  The effects of structural or observational uncertainty could be evaluated as an extension of the analyses presented above. For example, the sensitivity of \ac{do} to variation in the half-saturation constants for phytoplankton (the concentration supporting half the maximum uptake rate of nutrients) could vary given the initial nutrient concentrations \citep{Eppley69}. Further, changes in the ratio between nitrogen and phosphorus could affect the sensitivity of state variables to parameter changes depending on the limiting nutrient.  A more challenging analysis is an evaluation of the effects of structural components on model output, which requires exclusion or inclusion of explicit biogeochemical equations. Fortunately, the FishTank model includes several `switches' that allow users to change the governing equations that estimate state variables, such as switches that `turn on' different structural equations for light attenuation in the water column.  This design is uncommon in biogeochemical models and could be leveraged for an evaluation of structural uncertainty. As such, our analysis of parameter sensitivity and identifiabilty could be combined with an evaluation of observational and structural uncertainty for a more complete characterization of the model, having implications for understanding drivers of hypoxia in coastal waters.

>>>>

\title{Parameter sensitivity and identifiability for a biogeochemical model of hypoxia in the northern {G}ulf of {M}exico\tnoteref{mytitlenote}}
\tnotetext[mytitlenote]{Version: Mon Jul 31 17:43:04 2017 -0500, \href{https://github.com/fawda123/identifiability/commit/03ef21c36dd48fe1a6afd2f825a9bd6aaa19f078}{03ef21c36dd48fe1a6afd2f825a9bd6aaa19f078}\newline Acronyms: \acfi{chla}; \acfi{cgem}; \acfi{do}; \acfi{dom}; \acfi{gom}; \acfi{lcs}; \acfi{marb}; \acfi{pom}; \acfi{rmse}; \acfi{zerod}}

@@

\begin{abstract}
\noindent Local sensitivity analyses and identifiable parameter subsets were used to describe numerical constraints of a hypoxia model for bottom waters of the northern Gulf of Mexico.  The sensitivity of state variables differed considerably with parameter changes, although most variables were responsive to changes in parameters that influenced planktonic growth rates and less sensitive to physical or chemical parameters.  Variation in sensitivity had a direct correspondence with identifiability, such that only small subsets of the complete parameter set had unique effects on the model output. Selecting parameters by decreasing sensitivity demonstrated that only eight of 51 total parameters had a sufficiently unique effect on model output for accurate calibration.  As a result, parameter selection heuristics were used to identify parameters for model calibration that depended on combined effects on output, relative sensitivity of each parameter, and ecological categories for the biogeochemical equations. The calibrated \ac{zerod} unit of the hypoxia model had improved fit to the observed data if sensitive parameters were included in an identifiable subset.  Extension of results to a three-dimensional grid of the Gulf of Mexico showed that sensitive parameters for the \ac{zerod} model translated to non-trivial changes in the areal estimates of hypoxia.
\end{abstract}

@@

Calibration of parameter values for each of the seven selection heuristics (\cref{tab:heurist1,tab:heurist2,tab:heurist3})  produced model output that varied in goodness of fit and computational effort required to identify optimal values (\cref{tab:calib}).  As expected, the number of required iterations of the optimization algorithm to minimize \ac{rmse} of observed \ac{do} and model output varied in proportion to the number of parameters in a subset.  For example, only six iterations were required to optimize the single parameter selected for the temperature category, whereas 1530 iterations were required for the eight parameters selected independent of category.  No association was observed between collinearity ($\gamma$) and number of iterations or calibration.  Of the seven parameter subsets, selection independent of parameter category and equally within each category produced the largest reduction in \ac{rmse} values from the default output (starting \ac{rmse} $72.14$ mmol O$_2$ m$^{-3}$, reduction to $66.77$, $67.44$, respectively), whereas the optimal values for parameters in the Optics category had the smallest reduction ($72.12$).  Lower \ac{rmse} was associated with higher average sensitivity of a parameter subset.  Parameters selected only in the optics category had the highest \ac{rmse} and lowest average L1 ($2.81\times 10^{-4}$, three parameters), whereas parameters selected independent of category had the lowest \ac{rmse} and highest average L1 ($0.03$, eight parameters, \cref{tab:calib}).

@@

State variables were most sensitive to phytoplankton and zooplankton parameters, particularly the maximum growth rates (\textit{umax} for phytoplankton, \textit{Zumax} for zooplankton), mortality coefficient for phytoplankton (\textit{mA}), and the zooplankton half saturation coefficent for grazing (\textit{ZKa}). An increase in the growth rate of primary producers can increase oxygen concentration through photosynthetic processes, although increased production of organic matter is balanced with respiration and bacterial decomposition that reduce \ac{do} in the water column.  Similarly, increases in zooplankton abundance with increased growth rates can reduce phytoplankton biomass through grazing, which is expected to further deplete pools of organic matter from algal sources.  Most variables were also sensitive to variation in the half-saturation grazing coefficient which moderates nutrient concentrations that support half the maximum grazing rate. Although the tradeoff between abundance, grazing, and decomposition is complex, the sensitivity of model state variables to parameters that directly control the abundance of primary producers is in agreement with empirical observations of factors that influence hypoxia dynamics on the \ac{lcs} \citep{Fahnenstiel95,Roelke00,Eldridge10}. The sensitivity of the model output to variation in other parameters that relate to physical and chemical properties of the system was of secondary importance.  That is, state variables were sensitive to changes in light and temperature parameters, although to a lesser extent than biological parameters.  As such, the differing sensitivities of state variables to parameters in each of the categories was not unexpected given ecological relationships that are well understood and described by the model.

A general conclusion from the identifiability analyses is that only limited subsets of parameters were identifiable within the constraints of local sensitivity analyses. These results support previous studies that have suggested similarly small subsets of parameters can be identified using traditional calibration schemes \citep[e.g.,][]{Wheater86,Ye97,Omlin01}.  In addition to \ac{cgem}, these conclusions have relevance for other biogeochemical models that include numerous parameters and structural equations to characterize processes in the model domain.  A general conclusion is that less complex models are potentially beneficial given that only a small subset of parameters is identifiable and that ecosystem processes may in fact be sufficiently characterized with few parameters \citep{Ye97}.  Conversely, others have argued that model complexity is not in itself a disadvantage when parsimony is not the only determinant of model structure \citep{Reichert97}. Over-parameterization can be useful if processes have importance that were not evaluated during model identification.  Single objective functions that maximize model fit with identifiable parameters may also provide an incomplete characterization of model worth, which has prompted the development of probability-based models of hypoxia that explicitly include uncertainty in model components \citep[e.g.][]{Obenour15}.  Our results demonstrated that approximately 75\% of the evaluated parameters had an effect on the eight state variables, whereas \ac{cgem} includes a total of 36 variables and multiple plankton groups, not all of which have immediate concern for understanding hypoxia. The redundancies identified with the sensitivity analyses are challenging only if the primary interest is, for example, \ac{do} dynamics.  Moreover, the proposed selection heuristics provide flexibility for choosing different parameters with the assumption that those chosen depend on the research or management question.

Results from the identifiability analyses provided additional insight into the interactions of parameters in large biogeochemical models.  First, identifiability of parameter subsets was not related to the sensitivity of individual variables. As noted above, an identifiable parameter is one that has a unique effect on model predictions that cannot be compensated for or undone by changing other parameters. The magnitude of the effect of a parameter has no bearing on identifiability, which further complicates the selection of parameters for calibration.  Although identifiability is the primary limiting factor in choosing a set, the relative sensitivities are more important for the decision to include or exclude individual parameters.  For example, \cref{tab:calib,fig:areachg} demonstrated that parameter sets with sensitive parameters had improved model fit and that these parameters had affects that translated to the larger model. Our analysis addressed this challenge by presenting multiple selection criteria for identifiable parameter sets that prioritized the most sensitive parameters during the selection process.  Similarly, identifiability was not always related to the number of parameters in a set. Although the general trend was decreasing identifiability with more parameters, the unique effects of including an individual parameter with an existing set often reduced the $\gamma$ estimate. For example, \cref{fig:exclex} showed that including \textit{KGcdom}, \textit{KO2}, or \textit{nitmax} in parameter sets more often reduced $\gamma$ relative to sets that excluded the parameters.  The selection criteria proposed above can facilitate parameter selection and also provide diagnostic tools to identify parameters with disproportionate effects on $\gamma$.

@@

Our results demonstrated that small parameter subsets relative to all sensitive parameters were within the identifiability thresholds described in the literature.  The identifiable parameter subsets varied considerably between state variables and the method for parameter selection.  Use of identifiable parameter subsets with sensitive parameters improved model fit and results for the smaller model were scalable to the 3-dimensional model.  However, an evaluation of sensitivity and identifiability of relevant parameter sets is a preliminary and simplistic approach to improving model predictions.  We have provided a general approach to select parameter subsets for further model development depending on the ecological context (i.e., selection by parameter category, selection for specific state variables). Thus, the results described above have relevance for model refinement with the specific goal of better understanding ecological dynamics that moderate hypoxia on the northern \ac{gom}.  The general principles of sensitivity and parameter identifiability have broad applicability beyond this context and we argue that such methods should be more universally applied as an initial approach to quantify numerical constraints of biogeochemical models.

The effects of structural or observational uncertainty could be evaluated as an extension of the analyses presented above. For example, the sensitivity of \ac{do} to variation in the half-saturation constants for phytoplankton (the concentration supporting half the maximum uptake rate of nutrients) could vary given the initial nutrient concentrations \citep{Eppley69}. Further, changes in the ratio between nitrogen and phosphorus could affect the sensitivity of state variables to parameter changes depending on the limiting nutrient.  A more challenging analysis is an evaluation of the effects of structural components on model output, which requires exclusion or inclusion of explicit biogeochemical equations. The FishTank model includes several `switches' that allow users to change the governing equations that estimate state variables, such as switches that `turn on' different structural equations for light attenuation in the water column.  This design is uncommon in biogeochemical models and could be leveraged for an evaluation of structural uncertainty. As such, our analysis of parameter sensitivity and identifiabilty could be combined with an evaluation of observational and structural uncertainty for a more complete characterization of the model, having implications for understanding drivers of hypoxia in coastal waters.
